Differential Protein expression

### 2021-10-28 reports/rmarkdown/manuscript/Figure8.Rmd
Figure8. with cytokines

```{r message=FALSE, warning=FALSE, include=FALSE}
##################
# LOAD LIBRARIES #
##################
#setwd("/Users/vilkal/work/Brolidens_work/Projects/Gabriella_repo/reports/rmarkdown/manuscript")
suppressWarnings({suppressMessages({suppressPackageStartupMessages({
  library(tidyverse)
  library(edgeR)
  library(openxlsx)
})  })  })

#############
# LODA DATA #
#############
Protein_V3 <- read_csv("../../../data/2020_04_20_LAMIQ_visit_3_singlets.csv") 
datasets_all_samples <- readRDS("../../../results/datasets_all_samples.RDS")
metadata <- read.csv2("../../../results/metadata_integration.csv",row.names = 1, stringsAsFactors = F)

@@

# top_PROT_mfi <- t
# gr <- "Luminal_gr"
# plot_heatmap.fun(t, "Luminal_gr", scale = T)
plot_heatmap.fun <- function(top_PROT_mfi, gr, scale = FALSE){
  df <- na.omit(metadata[sample_use, c(gr, "ID")])
  df <- arrange(df, df[1])
  samples <- df[, "ID"]
  names(samples) <- df[, gr]

@@

  layout(matrix(c(1,2,3,4,
                  1,2,3,4),
                  nrow = 1,ncol = 2,byrow = T),widths = c(1.,2,.4,6) )
    ## Heatmap
    par(mgp=c(0,0.2,0), mar=c(4,.4,1,0)) #(bottom, left, top, right) 
    plot( rev(as.dendrogram(h)) , xlim=c(max(h$height),-.01),
          horiz = T, leaflab = "none",xaxs="i",yaxs="i",axes=T, cex.axis = 0.7 )
    #title(xlab="height", line=3)
    mtext( c("Height"), side=1, las = 1, cex = 0.7, line =1,xpd=T)

    if(gr == "Tissue_gr") {
      height <-1.5 }else{height <- round(max(h$height))}
    lines(x=c(0,height), y=c(0.45,0.45), col="black", lty=1, lwd = 2.5) # only for grob

    # abline(v=cuttoff,xpd=F,col="red",lty=2)
    # points( rep(-.1,length(gene_module)),
    #         seq(length(gene_module),1,length.out = length(gene_module)),
    #         col=taxa_pal[factor(gene_module[h$order])],
    #         pch=15,cex=.5,xpd=F)

    par(mar=c(4,.2,1,2)) #(bottom, left, top, right) 
    image( t(top_PROT_mfi[h$order,pull(df, "ID")][nrow(top_PROT_mfi):1,]),
           col = colorRampPalette(c("navy","navy","grey95","firebrick4","firebrick4") )(99),
           breaks = breaks,
           axes=F)

    mtext( c("patient groups"), side=1, las = 1, cex = 0.7, line =1,xpd=T)

@@

### top dpe with p-value < 0.05 un-scaled and scaled respectively
```{r single plots, fig.height=5, fig.width=3, include=FALSE}
# top differentially expressed proteins Luminal groups
t <- top_dpe.fun(dpe_L$top)

library(cowplot)
plot_heatmap.fun(t, "Luminal_gr")
p.plot <- recordPlot()
plot_heatmap.fun(t, "Luminal_gr", scale = T)
p.plot_s <- recordPlot()

# top differentially expressed proteins Tissue groups
t <- top_dpe.fun(dpe_T$top)

library(cowplot)
plot_heatmap.fun(t, "Tissue_gr")
p.plot_T <- recordPlot()
plot_heatmap.fun(t, "Tissue_gr", scale = T)
p.plot_s_T <- recordPlot()

@@

```{r top dpe Luminal plot, echo=FALSE, fig.height=5, fig.width=6}
# side by side
# plot_grid(ggdraw(p.plot), ggdraw(p.plot_s), 
#           labels = c('a', 'b'),rel_widths=c(1,1), greedy = F)

plot_grid(ggdraw(p.plot_s),ggdraw(p.plot_s_T), 
          labels = c('a', 'b'),rel_widths=c(1,1), greedy = F) # , 'c', 'd'

@@

```{r top dpe Tissue plot, eval=FALSE, fig.height=10, fig.width=6, include=FALSE}
# side by side Luminal & Tissue
# plot_grid(ggdraw(p.plot), ggdraw(p.plot_s),
#           ggdraw(p.plot_T), ggdraw(p.plot_s_T), 
#           labels = c('a', 'b', 'c', 'd'),rel_widths=c(1,1), greedy = F)

plot_grid(ggdraw(p.plot_s),ggdraw(p.plot_s_T), 
          labels = c('a', 'b'),rel_widths=c(1,1), greedy = F) # , 'c', 'd'

@@

**Figure 3. Protein Heatmap**. **a** and **c** Without scaling (normalized MFI), for Luminal and Tissue respectively **b** and **d** scaled by row for Luminal and Tissue respectively

## all dpe un-scaled and scaled respectively
### Luminal
```{r all dpe plot, eval=FALSE, fig.height=6, fig.width=3, include=FALSE}
# all proteins 
plot_heatmap.fun(Protein[,sample_use], "Luminal_gr")
plot_heatmap.fun(Protein[,sample_use], "Luminal_gr", scale = T)
```

### Tissue
```{r all dpe Tissue plot, eval=FALSE, fig.height=6, fig.width=3, include=FALSE}
# all proteins 
plot_heatmap.fun(Protein[,sample_use], "Tissue_gr")
plot_heatmap.fun(Protein[,sample_use], "Tissue_gr", scale = T)
```

>>>>

```{r message=FALSE, warning=FALSE, include=FALSE}
##################
# LOAD LIBRARIES #
##################
#setwd("/Users/vilkal/work/Brolidens_work/Projects/Gabriella_repo/reports/rmarkdown/manuscript")
suppressWarnings({suppressMessages({suppressPackageStartupMessages({
  library(tidyverse)
  library(edgeR)
  library(openxlsx)
  library(rstatix)
  library(broom)
  library(niceRplots)
})  })  })

#############
# LODA DATA #
#############
Protein_V3 <- read_csv("../../../data/2020_04_20_LAMIQ_visit_3_singlets.csv") 
datasets_all_samples <- readRDS("../../../results/datasets_all_samples.RDS")
metadata <- read.csv2("../../../results/metadata_integration.csv",row.names = 1, stringsAsFactors = F)
cytokine_data <- read.csv("../../../data/210125_Cytokines_visit3CVL.csv")

@@

# top_PROT_mfi <- t
# gr <- "Luminal_gr"
# plot_heatmap.fun(t, "Luminal_gr", scale = T)
plot_heatmap.fun <- function(top_PROT_mfi, gr, scale = FALSE, layout = "none"){
  df <- na.omit(metadata[sample_use, c(gr, "ID")])
  df <- arrange(df, df[1])
  samples <- df[, "ID"]
  names(samples) <- df[, gr]

@@

  if(layout == "multi") {
    layout(matrix(c(1,2,5,6,
                    7,7,8,8,
                    3,3,4,4),
           nrow = 3, ncol = 4,byrow = T),
           widths = c(.9,2,.9,2), heights = c(1,1,.01) ) }
  if(layout == "single") {
    layout(matrix(c(1,2),
                    nrow = 1, ncol = 2,byrow = T),widths = c(1,2) ) }

    ## Dendogram
    par(mgp=c(0,0.2,0), mar=c(4,.6,1,.2)) #(bottom, left, top, right) 
    plot( rev(as.dendrogram(h)) , xlim=c(max(h$height),-.01),
          horiz = T, leaflab = "none",xaxs="i",yaxs="i",axes=T, cex.axis = .7 )
    #title(xlab="height", line=3)
    mtext( c("Height"), side=1, las = 1, cex = .7, line =1,xpd=T)

    if(gr == "Tissue_gr") {
      height <-1.5 
      add_letter("b")
    }else{height <- round(max(h$height))
      add_letter("a")}
    #lines(x=c(0,height), y=c(0.45,0.45), col="black", lty=1, lwd = 2.5) # only for grob

    ## Heatmap
    par(mar=c(4,.2,1,2.5)) #(bottom, left, top, right) 
    image( t(top_PROT_mfi[h$order,pull(df, "ID")][nrow(top_PROT_mfi):1,]),
           col = colorRampPalette(c("navy","navy","grey95","firebrick4","firebrick4") )(99),
           breaks = breaks,
           axes=F)

    mtext( c("patient groups"), side=1, las = 1, cex = .7, line =1,xpd=T)

@@

```{r sig. cytokines function, include=FALSE}
sig_stars.fun <- function(df_krusk, df_dunn, group, n = 6, order = NULL){
  samples_use <- metadata[,c(group, "ID")] %>% na.omit() %>% filter(ID %in% colnames(cytokine_data))

  # select only sig. cytokines 
  sig <- df_krusk %>% filter(p.adj<0.05) %>% pull(Cytokines)
  sig_df <- cytokine_data_log[sig,samples_use$ID] %>%
    cbind(Max = c(rep(max(.)+n, nrow(.))))

  groupings <- samples_use[[group]] [ match( colnames(sig_df[,1:ncol(sig_df)-1]) , as.character(samples_use$ID)  ) ]
  groupings <- c(groupings, "Max")

   if(is.null(order)){
    g <- rownames(sig_df)
  }else{g <- order}

  stars_df <- df_dunn %>%
    mutate(x = (as.integer(str_extract(.$group1, "\\d")) + 
                  as.integer(str_extract(.$group2, "\\d")))/2) %>%
    mutate(line_x0 = as.integer(str_extract(.$group1, "\\d"))) %>%
    mutate(line_x1 = as.integer(str_extract(.$group2, "\\d"))) %>%
    mutate(y = (as.numeric(factor(Cytokines, levels = rev(g))))-.3) %>%
    arrange(group1, group2) %>%
    dplyr::group_by(Cytokines) %>%
    nest() %>%
    mutate(y_ = map(data, ~accumulate(.$y, ~.x + .1))) %>%
    unnest(c(data, y_))

  return(list(sig_df=sig_df, stars_df=stars_df, groupings=groupings, samples_use=samples_use))
}
```

```{r Cytokine stats tests, include=FALSE}
cytokine_data$ID <- metadata [ match( cytokine_data$PatID , as.character(metadata$PatID)  ) ,"ID" ]
cytokine_data <- cytokine_data[ !is.na(cytokine_data$ID) , ]
rownames(cytokine_data) <- cytokine_data$ID
cytokine_data <- cytokine_data[,grepl("CVL",colnames(cytokine_data))]
colnames(cytokine_data) <- sub("CVL_","",colnames(cytokine_data))
cytokine_data <- t(cytokine_data)
cytokine_data[is.na(cytokine_data)] <- 0.01
cytokine_data <- log2(cytokine_data+1)
cytokine_data_log <- cytokine_data

cyto2 <- as.data.frame(t(cytokine_data))
cyto2$ID<-row.names(cyto2)

#select data and re-format to long format
cytogr.long <- metadata %>%
  select(ID, Luminal_gr, Tissue_gr) %>%
  left_join(.,cyto2, by="ID") %>%
  pivot_longer(c(-Luminal_gr, -Tissue_gr, -ID), names_to = "Cytokines", values_to = "value")

test.krusk_dunn.fun <- function(df, var, group) {
  var <- enquo(var)

  stat.test.krusk <- df %>%
  group_by(!!(var)) %>%
  kruskal_test(., as.formula(paste("value", "~", group))) %>%
  adjust_pvalue(method = "BH") %>%
  select(-.y., -n, -df) %>%
  arrange(p) %>%
  rename(p.value = "p") %>%
  add_significance()

# select the significant proteins from kruskal
sig <- stat.test.krusk %>% filter(p.adj<0.05) %>% pull(!!var)

### Post hoc test
stat.test.dunn <- df %>%
  filter(!!var %in% sig) %>%
  group_by(!!var) %>%
  dunn_test(., as.formula(paste("value", "~", group))) %>%
  #dunn_test(value ~ !!group) %>%
  select(-.y.) %>%
  arrange(p) %>%
  rename(p.value = "p") %>%
  filter(p.adj<0.05)

  sig2 <- stat.test.dunn[stat.test.dunn$p.adj<0.05,]

 l <- list(stat.test.krusk, stat.test.dunn)
 n <-  c("Kruskal", "Dunn's") %>% paste0(., "_", group, sep = "")
 l <- set_names(l, n)

return(l)
}

stat <- imap(c("Luminal_gr", "Tissue_gr"), ~test.krusk_dunn.fun(cytogr.long, Cytokines, .x))
s <- flatten(stat)
write.xlsx(s, file=paste0("../../../results/","Cytokine_stats",".xlsx"))

@@

```{r Proteins plot, echo = FALSE, eval = TRUE, fig.height=10, fig.width=7}
figlabels <- letters

# top differentially expressed proteins Luminal groups
t <- top_dpe.fun(dpe_L$top)
plot_heatmap.fun(t, "Luminal_gr", scale = T, layout = "multi")

# top differentially expressed proteins Tissue groups
t <- top_dpe.fun(dpe_T$top)
par(mfg = c(2,1)) # adds the remaning plots to the previous layout
plot_heatmap.fun(t, "Tissue_gr", scale = T)

# Luminal Cytokine plot 
g <- c("IL1a", "MIG", "IP10", "IL1b", "IL12_p70", "MCP1")
Luminal <- sig_stars.fun(s$Kruskal_Luminal_gr, s$`Dunn's_Luminal_gr`, "Luminal_gr", order = g)
temp <- as.factor(c(Luminal$groupings, "Max"))
stars <- Luminal$stars_df

par( mar=c(1.4,4,1.5,1)) #b,l,t,r 
violist(data=Luminal$sig_df,
          srt=45,
          genes= g,
          col = pal, transparency = 50,
          clustering=c(Luminal$groupings),smooth = 2 )
segments(x0 = stars$line_x0, x1 = stars$line_x1, y0 = stars$y_, y1 = stars$y_) 
text( stars$x , stars$y_+.05,
        labels = stars$p.adj.signif, srt = 0, xpd = TRUE, cex=1)
title(main = "Luminal gr. Cytokines", line = 0.7, cex.main = .9)

add_letter("c")

# Tissue Cytokine plot
Tissue <- sig_stars.fun(s$Kruskal_Tissue_gr, s$`Dunn's_Tissue_gr`, "Tissue_gr")
temp <- factor(Tissue$groupings, levels=c("T1","T2","T3","T4","Max"))
stars <- Tissue$stars_df

par( mar=c(24.5,5,1.5,1)) #b,l,t,r
violist(data=Tissue$sig_df,
          srt=45, 
          genes= rownames(Tissue$sig_df),
          col = pal, transparency = 50,
          clustering=Tissue$groupings, smooth = 2 )
segments(x0 = stars$line_x0, x1 = stars$line_x1, y0 = stars$y_, y1 = stars$y_) 
text( stars$x , stars$y_+.05,
        labels = stars$p.adj.signif, srt = 0, xpd = TRUE, cex=1)
title(main = "Tissue gr. Cytokines", line = 0.7, cex.main = .9)

add_letter("d")

@@

**Figure 3. Protein Heatmap**. **a** and **c** Without scaling (normalized MFI), for Luminal and Tissue respectively **b** and **d** scaled by row for Luminal and Tissue respectively. top dpe with p-value < 0.05

```{r All dpe plot, eval=FALSE, fig.height=6, fig.width=6, include=FALSE}
library(cowplot)

plot_heatmap.fun(Protein[,sample_use], "Luminal_gr")
p.plot <- recordPlot()

plot_heatmap.fun(Protein[,sample_use], "Tissue_gr")
p.plot_T <- recordPlot()

# side by side unscaled
plot_grid(ggdraw(p.plot),ggdraw(p.plot_T), 
          labels = c('a', 'b'),rel_widths=c(1,1), greedy = F) # , 'c', 'd'

@@
